---
title: "Stats camp day 1"
author: "Frank Edwards"
date: "8/28/2019"
output: binb::metropolis
---

# Welcome to Graduate statistics training!

## Stats camp goals

1. Get your software set up
2. Remind you what these things mean:

 \[log(x)\]
 \[y = 2 + 3x\]
 \[\frac{d}{dx}x^3  = 3x^2 \]
 \[\sum_{x=1}^{10} x \]

## Today: Let's install software

**Step 1**

Install R. https://cran.r-project.org/

**Step 2**

Install RStudio. https://www.rstudio.com/

\pause

- Does anyone have problems with regular access to a computer that can run R?

# Problems?

## Getting started in RStudio

- The script and the console \pause
- Make it pretty! (You will be spending lots of time here...)
  - Themes
  - Pane layouts

## Getting started in RStudio: packages

- Install packages

```{r eval = FALSE, size = "tiny"}
install.packages("tidyverse")
```

## Getting started in RStudio: packages

- Load packages

```{r}
library(tidyverse)
```

# Problems?

## Using R as a (very fancy) calculator

```{r eval = F}
2 + 2
```

## Using R as a (very fancy) calculator

```{r eval = F}
2 * 2
```

## Using R as a (very fancy) calculator

```{r eval = F}
2 / 2
```


## Using R as a (very fancy) calculator

```{r eval = F}
2 ^ 3
```

## Using R as a (very fancy) calculator

```{r eval = F}
sqrt(4)
```

## Using R as a (very fancy) calculator

```{r eval = F}
# I am a comment! I help you read code!
# the <- operator makes assignments
# Make a new variable called 'x', set it equal to 2
x <- 2
x^2
```

## Using R as a (very fancy) calculator

Compute the following in R (with $x=2$)

- $x^3$ \pause
- $2x$ \pause
- $\frac{x}{2}$ \pause
- $(2+x)^2$ \pause

## Before we proceed

- Are you running RStudio? \pause
- Have you installed tidyverse? \pause
- When you type 2==3 in the console and hit enter, what does it say? \pause

# Exciting! You are programming in R!
 
# Break
 
# Welcome back: math time!

## Agenda today

1. Math notation and key concepts
2. Exponents, logarithms
3. Lines and graphs

## Real numbers and integers

**Real numbers (doubles)**

- Any continuous number
- E.g. 4, 4.189, 2/3, $\pi$

**Integers**

- Any whole number
- 10, -10, 24, 87

## Variables

- May take on any value
- Represented by letters, such as $x, y, z$
- Can be used in any mathematical operation

## Functions

- Maps each element of set $x$ to an element of set $y$
- Often denoted by $f, g, h$
- $f(x) = 2x + 3$

## Parameters (statistics)

- Variable that represent a feature of a population
- Represented by Greek letters, such as $\mu, \sigma, \varepsilon$

## Summation

Represented as $\sum$, with integer begin and end points 

$$\sum_{x=1}^{3} x$$
\pause 

$$\sum_{x=1}^{3} x = 1 + 2 + 3 = 6$$

## Summation in R

In R, we can calculate a sum using the sum() function

```{r}
# make an integer vector from 1 to 3
x<-1:3
# x<-c(1,2,3) is equivalent
sum(x)
```

## Summation review

Compute the following by hand, and then in R

$$\sum_{x=3}^8 (x+1)$$

$$\sum_{x=1}^4 2x $$

## Exercises (solutions)

Compute the following in R

$$\sum_{x=3}^8 (x+1)$$

$$\sum_{x=1}^4 2x $$

## Exercises (solutions)

Compute the following in R

$$\sum_{x=3}^8 (x+1)$$
```{r eval = F}
x<-3:8 # or c(3, 4, 5, 6, 7, 8)
sum(x+1)
```



$$\sum_{x=1}^4 2x $$

```{r eval = F}
x<-c(1, 2, 3, 4) # or 1:4
sum(2*x)
```

## Products

Represented as $\prod$, with integer begin and end points 

$$\prod_{x=1}^4 x  $$

\pause

$$\prod_{x=1}^4 x = 1 \times 2 \times 3 \times 4 $$

\pause

In R: 

```{r}
x<-1:4
prod(x)
```

# Exponents and logarithms

## Exponents

$a^n$ Is equal to a multiplied by itself n times \pause

- $2^3 = 2 \times 2 \times2 = 8$ \pause
- $5^4 = 5 \times 5 \times 5 \times 5 = 625$

In R, we can calculate exponents using the ^

```{r}
2^3
5^4
```

## Some rules of exponents

$$x^1 = x$$ \pause
$$x^0 = 1$$ \pause
$$x^k + x^l = x^{k+l}$$ \pause
$$(x^k)^l = x^{kl}$$ \pause
$$(xy)^k = x^k \cdot y^k$$ \pause

## More rules of exponents

$$ \left( \frac{x}{y} \right) ^k = \left( \frac{x^k}{y^k} \right) $$ \pause
$$\frac{x^k}{x^l} = x^{k-l}$$ \pause
$$x^{1/k} = \sqrt[k]{x}$$ \pause 
$$x^{-k} = \frac{1}{x^k}$$ \pause 

## Exponents in R

```{r}
x<-4
x^3
```
\pause
```{r}
x^(2+3)
```
\pause
```{r}
## for base e, use exp()
exp(4)
```





## Logarithms

A logarithm is the power $x$ required to raise a base $c$ to a given number $a$. \pause

$$\log_c(a) = x$$
$$c^x = a$$ \pause
$$3^2=9$$ \pause
$$\log_3(9) = 2$$ 

## Common logarithms

- The most common log bases are 2, 10, and $e = 2.718$
- Log with base $e$ is called a natural log, $ln$
- The R function log() has a default base $e$
- We use log base $e$ to model many exponential growth processes 

## Logarithms are the inverse of exponents

$$10^2=100$$ \pause
$$\log_{10}(100)=2$$ \pause
$$e^2 = 7.389056$$ \pause
$$\log_e(7.389056) = 2 $$ 

## Logarithm rules

$$\log(x \cdot y) = \log(x) + \log(y)$$ \pause
$$log(x^n)  = n \log(x)$$ \pause
$$\log\left(\frac{x}{y}\right) = \log(x) - \log(y)$$

## Logarithms in R

For statistics, it is safe to assume that $log$ means $ln$. In R, this is the default

```{r}
log(10)
log(10, base = 10)
```

## Exercises

Compute the following in R. Assume $x=6$

1. $x^4$ 
2. $2x^{5+x}$
3. $e^x$
4. $log(x)$
5. $log(2x + 3)$
6. $log(\frac{1}{2x})$

# Coordinates and lines

## The coordinate plane

```{r echo = FALSE, message = FALSE, fig.width = 4, fig.height = 2.5}
library(ggplot2)
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```

## Plotting points

For coordinate pairs $(x_1,y_1), (x_2, y_2), (x_3, y_3)$, we can plot each point along an x and y axis.

**Example:**  $(0,-2), (4,3), -6, 8)$

```{r echo = FALSE, message = FALSE, fig.height = 2.5, fig.width = 4}
dat<-data.frame("x" = c(0, 4, -6), "y" = c(-2, 3, 8))
ggplot(dat, aes(x = x, y = y)) + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
  
```

## Lines

The typical equation for a line is $y = mx + b$ where m is the slope and by is the y-intercept. \pause

You will often see a line expressed as a regression equation: 

$$y = \beta_0 + \beta_1x$$  

where $\beta_0$ is the y-intercept and $\beta_1$ is the slope.

## The slope

Slope measures the steepness of a line. A line with a positive slope has increasing values of y as x increases. A line with a negative slope has decreasing values of y for increasing values of x. \pause

We can calculate the slope with two coordinates on the line $(x_1, y_1), (x_2, y_2)$ \pause

The slope is the ratio of the difference in y values to the difference in x values.

$$ m = \frac{y_2-y_1}{x_2-x_1}$$

## The y-intercept

The y-intercept is the value of y when x = 0. If we have the value of one point on the line, and the slope, we can obtain the y-intercept \pause

$$ b = y_1 - m \cdot x_1$$

## Example: intercept = 0, slope = 1

```{r echo = F}
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_abline(slope = 1, intercept = 0, lty = 2) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```


## Example: intercept = 0, slope = 0.2

```{r echo = F}
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_abline(slope = 0.2, intercept = 0, lty = 2) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```

## Example: intercept = 0, slope = -1

```{r echo = F}
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_abline(slope = -1, intercept = 0, lty = 2) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```

## Example: intercept = 5, slope = -1

```{r echo = F}
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_abline(slope = -1, intercept = 5, lty = 2) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```

## Example: intercept = -5, slope = -1

```{r echo = F}
ggplot() + 
  geom_point() + 
  xlim(-10,10) + 
  ylim(-10,10) +
  ylab("y") + 
  xlab("x") + 
  geom_abline(slope = -1, intercept = -5, lty = 2) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
```

## Example

Given the points $(1,3)$ and $(4,6)$ , the slope is $m = \frac{6-3}{4-1} = 1$ and the y-intercept is $b = 3 - 1 \cdot 1 = 2$. The equation of the line is $y= 1x + 5$

```{r echo = FALSE, fig.height = 2.5, fig.width = 4}
dat<-data.frame(x = c(4, 1), y = c(6,3))
ggplot(dat) + 
  geom_abline(intercept = 2, slope = 1, lty = 2) +
  geom_point(aes(x = x, y = y)) +
  xlim(-10,10) + 
  ylim(-10,10) + 
  ylab("y") + 
  xlab("x") + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
  
```

# Functions

## Functions, in general

A function maps each element in a set X to an element in set Y

- Linear function: $f(x) = x + 5$ \pause
- Quadratic function: $f(x) = x^2 + 2x + 3$ \pause
- Exponential function: $f(x) = e^{2x} + 6$

## Graphical forms of functions: linear

$f(x) = 2x + 5$

```{r echo = FALSE, fig.height = 2.5, fig.width = 4}
dat<-data.frame(x = c(4, 1), y = c(6,3))
ggplot(dat) + 
  geom_abline(intercept = 5, slope = 2, lty = 2) +
  xlim(-10,10) + 
  ylim(-10,10) + 
  ylab("y") + 
  xlab("x") + 
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()
  
```

## Graphical forms of functions: quadratic

$f(x) = x^2 + 2x + 3$

```{r echo = FALSE}
quad<-function(x){x^2 + 2*x + 3}
ggplot(data.frame(x = c(-10,10)), aes(x=x)) + 
  stat_function(fun = quad) + 
  theme_bw()
```

## Graphical forms of functions: exponential

$f(x) = e^{2x} + 6$

```{r echo = FALSE}
exp_fun<-function(x){exp(2*x) + 6}
ggplot(data.frame(x = c(0,4)), aes(x=x)) + 
  stat_function(fun = exp_fun) + 
  xlim(c(0, 5)) +
  theme_bw()
```

## Graphical forms of functions: logarithmic

$f(x) = log(x)$

```{r echo = FALSE}
log_fun<-function(x){log(x)}
ggplot(data.frame(x = c(0,4)), aes(x=x)) + 
  xlim(c(1, 100)) + 
  stat_function(fun = log_fun) + 
  theme_bw()
```

## Functions in R 

We can easily define functions in R. 

$$f(x) = 2(x + 3)^3$$

\pause

```{r}
f_x<-function(x){2 * (x + 3)^2}
f_x(3)
```

\pause

$$g(x) = \frac{x-1}{5} $$

\pause

```{r}
g_x<-function(x){(x-1)/5}
g_x(4)
```

## Exercises

Define and evaluate the following functions in R. Assume $x=2$

1. $f(x) = 2x$
2. $f(x) = \frac{x}{2}$
3. $f(x) = 2(x+1)^3$

