---
title: "Stats camp day 1"
author: "Frank Edwards"
date: "8/28/2019"
output: binb::metropolis
---

# Welcome to PhD statistics training!

## Stats camp goals

1. Get your software set up
2. Remind you what these things mean:

 \[log(x)\]
 \[y = 2 + 3x\]
 \[\frac{d}{dx}x^3  = 3x^2 \]
 \[\sum_{x=1}^{10} x \]

## Today: Let's install software

**Step 1**

Install R. https://cran.r-project.org/

**Step 2**

Install RStudio. https://www.rstudio.com/

# Problems?

## Getting started in RStudio

- Make it pretty! (You will be spending lots of time here...)
  - Themes
  - Pane layouts

## Getting started in RStudio: packages

- Install packages (tidyverse, devtools, swirl)
- Install QSS package: 

```{r eval = FALSE}
install.packages(c("tidyverse", "devtools", "swirl"))
library(devtools)
install_github("kosukeimai/qss-package", build_vignettes = TRUE)
library(swirl)
install_course_github("kosukeimai", "qss-swirl")
```

# Problems?

## Getting communication set up

- Slack
  - Invites
  - Tour

## Setting up Git

- Install Git: https://git-scm.com/downloads
- (Optional) Register for a GitHub account: https://github.com/
- Clone course repository

## Cloning the course gets you slides

First, choose a folder you want intro_stats to live in

```{r eval = FALSE}
git clone https://github.com/f-edwards/intro_stats.git
```

Then, to get updates as I push them up to the server

```{r eval = FALSE}
cd intro_stats
git pull
```

# Problems?

# A very brief introduction to markdown

## Why markdown?

1. It is pretty
2. It makes your workflow simpler
3. It is reproducible

# Let's play with it