---
title: "Importing data"
author: "Frank Edwards"
output: binb::metropolis
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(MASS)
select<-dplyr::select
set.seed(1)

options(xtable.comment = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "small")
```

# Pull ACS Data from IPUMS

## Setting up your data pull

- Create an account
- Select the most recent ACS (1-year) data for your sample
- Select the following person-level variables: Age; Race; Hispanic ethnicity; Educational attainment; Employment status; poverty status; Wage and salary income; Interest dividend, and rental income 
- Select the following household-level variables: Total Household Income
- Create data extract (format as comma separated .csv)
- Move to project directory

## Extract data 

- (optional for windows) Install Git for Windows
  - Set git bash as default terminal in tools-> global options -> terminal
  - or use 7zip or other archive utility
- on terminal `gunzip usa_001.tar.gz`
- create `./data/` and move the .csv file there

## File paths

- Use `getwd()` to identify your file path
- Use the file browser to locate your working directory in the GUI
- Paths and relative paths: Note the differences between `.` and `~`

## Import data

We can use `read` functions to import data. Here we will use `read_csv`. For now, just give it one argument, the file path, and assign it to an object
 
```{r}
acs<-read_csv("./data/usa_00082.csv")
```

# Commence analysis!


