---
title: "Linear Regression part 2"
author: "Frank Edwards"
output: binb::metropolis
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(qss)
library(MASS)
select<-dplyr::select
set.seed(1)

options(xtable.comment = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "small")
```

## Linear regression with multiple predictors

We can extend the linear regression model:

\[ y = \beta_0 + \beta_1X + \varepsilon\] \pause

to include more than one predictor. We rewrite the equation as:

\[ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 \cdots \beta_p x_p + \varepsilon\] \pause

## In matrix notation

To be more compact:

\[ Y = \beta X + \varepsilon\]

Where $Y$ is the vector of predictors, $\beta$ is the vector of coefficients (including the intercept), $X$ is the matrix of all predictors, and $\varepsilon$ is the error term.

## Estimating a multiple linear regression in R

```{r}
library(gapminder)
gm_model<-lm(lifeExp ~log(gdpPercap) + continent, 
               data = gapminder)
```

## Interpreting a regression model with multiple coefficients

```{r size = "tiny"}
broom::tidy(gm_model)
```

## Interpreting a regression model with multiple coefficients

```{r size = "tiny"}
broom::tidy(gm_model)
```

Recall that $\hat{y} = \beta_0 + \beta_1 x_1 + \beta_2 x_2$

What is our expected value for life expectancy:

- if GDP per capita = 1000 and continent = Europe \pause
- if GDP per capita = 2000 and continent = Asia \pause
- if GDP per capita = 3000 and continent = Africa

## Visualizing the model

```{r echo = FALSE}
coefs<-coef(gm_model)

ggplot(gapminder,
       aes(x = log(gdpPercap), 
           y = lifeExp, color = continent)) + 
  geom_point(alpha = 0.5) + 
  geom_abline(aes(intercept = coefs[1], slope = coefs[2]), color ="#F8766D") + 
  geom_abline(aes(intercept = coefs[1] + coefs[3], slope = coefs[2]), color ="#7CAE00") + 
  geom_abline(aes(intercept = coefs[1] + coefs[4], slope = coefs[2]), color ="#00BFC4") + 
  geom_abline(aes(intercept = coefs[1] + coefs[5], slope = coefs[2]), color ="#C77CFF") 
```

## Let's try this with different data

```{r}
### load in the 'mtcars' data
data(mtcars)
head(mtcars)
```

## Variables in the mtcars data

- mpg	Miles/(US) gallon
-	cyl	Number of cylinders
-	disp	Displacement (cu.in.)
-	hp	Gross horsepower
-	drat	Rear axle ratio
-	wt	Weight (1000 lbs)
-	qsec	1/4 mile time
-	vs	Engine (0 = V-shaped, 1 = straight)
-	am	Transmission (0 = automatic, 1 = manual)
-	gear	Number of forward gears
-	carb	Number of carburetors

## Let's build a model for fuel efficiency with this data (theory!)

Our outcome of interest is `mpg`. What measured features of these cars do we think might be related to fuel efficiency?

## Start simple

$$ \textrm{E}[\textrm{mpg}_i] = \beta_0 + \beta_1 \textrm{hp}_i $$ \pause
$$\textrm{mpg}_i = \beta_0 + \beta_1 \textrm{hp}_i + \varepsilon $$
How would we estimate this model in R? 

## Now let's add complexity

$$ \textrm{E}[\textrm{mpg}_i] = \beta_0 + \beta_1 \textrm{hp}_i + \beta_2 \textrm{wt}_i $$ \pause

# Build up and interpret an MPG model

## Interactions

The prior model allowed each continent to have its own intercept fpr life expectancy. What if we thought the relationship of between GDP and life expectancy varied by continent? 

We can add *interaction terms* to our model to model processes where we believe the relationship between $y$ and $x_1$ is a function of $x_2$. 

\[ y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_1 x_2 + \varepsilon\]

## Estimating interactions in R

```{r size = "tiny"}
gm_model2<-lm(lifeExp ~ log(gdpPercap) * continent,
               data = gapminder)
```

### Interpreting an interaction model

```{r}
broom::tidy(gm_model2)
```

- What is the expected value for life expectancy in the Americas when GDP per capita is 1500? 
- In Asia when GDP per capita is 1500?

## Visualizing interactions

```{r echo = FALSE}
coefs<-coef(gm_model2)

ggplot(gapminder,
       aes(x = log(gdpPercap), y = lifeExp, color = continent)) + 
  geom_point(alpha = 0.25) + 
  geom_abline(aes(intercept = coefs[1], 
                  slope = coefs[2]), color ="#F8766D") + 
  geom_abline(aes(intercept = coefs[1] + coefs[3], 
                  slope = coefs[2] + coefs[7]), color ="#7CAE00") + 
  geom_abline(aes(intercept = coefs[1] + coefs[4], 
                  slope = coefs[2] + coefs[8]), color ="#00BFC4") + 
  geom_abline(aes(intercept = coefs[1] + coefs[5], 
                  slope = coefs[2] + coefs[9]), color ="#C77CFF")
```
