---
title: "1. Introducing data analysis in R"
author: "Frank Edwards"
date: "9/4/2019"
output: binb::metropolis
---

# Introduction to Statistics

# Overview of the course

## Course Goals

1. Introduce students to statistical computing through the R programming language
2. Develop data manipulation, exploration, and visualization skills
3. Introduce core concepts in probability and statistics

##  Lecture / lab / office hours

- Lecture introduces new materials and concepts 
- Lab (W 1:30-3:00, SCJ computer lab) provides us a chance to work through a problem set together
- Office hours (FE: Monday 10:00-2:00, CS: TBA) give you a chance to seek additional help with homework problems

## Weekly routine for the course

1. Do the assigned readings
2. Attend lecture
3. Practice skills in lab
4. Do homework
5. Take a break
6. return to 1 

## Books and readings

1. Imai, *Quantitative Social Science*

This book is the foundation of the course. It introduces core social science methods, programming and statistics.

2. Wickham and Grolemund, *R for Data Science*

This book (available free at r4ds.had.co.nz/) provides a comprehensive overview of programming in R using the tidyverse packages

**Supplemental materials**

Arnold, *Quantitative Social Science: The R Tidyverse code*

jrnold.github.io/qss-tidy/

This website provides a complete translation of the base R code in QSS into tidyverse syntax

## Homework

- I will assign a problem set at the end of each lecture
- Problem sets will be posted at github.com/f-edwards/intro_stats/
- Data is posted either on GitHub or available through the QSS package
- Problem sets are due at 12PM the Tuesday before each lecture
- Complete homework using RMarkdown 
- Provide adequate writing to explain and contextualize your responses / findings
- Email me (frank.edwards@rutgers.edu) the compiled RMarkdown output file and source code (.html and .rmd files) by the due date
- Everyone gets two free 3-day extensions throughout the semester, just let me know if you need to use it

## Course communication

- Course website: https://f-edwards.github.io/intro_stats/
- Course Slack: https://scj-introstats.slack.com

# Questions about the course?

## Software installation

- R: https://cran.r-project.org/
- RStudio: https://www.rstudio.com/
- Git (optional): https://git-scm.com/
- GitHub course repo: https://github.com/f-edwards/intro_stats

## Packages for the course

```{r message = FALSE, warning = FALSE, echo = TRUE}
#install.packages(c("devtools", "tidyverse"))
library(devtools)
### For book data
install_github("kosukeimai/qss-package", 
               build_vignettes = TRUE)
### For problem set data
install_github("conjugateprior/qss.student")
```

# Problems?

# Introducing: R!

## R is a calculator

```{r}
5+3
5*3
5/3
5^3
```

## R can make comparisons

```{r}
5>3
5<=3
5==3
"a"=="a"
```

## R works with objects 

```{r}
a<-2
a+1
b<-a+2
b
a<-a+1
a
```

## Objects can take many types

```{r}
a<-2
class(a)
b<-"howdy"
class(b)
c<-TRUE
class(c)
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector1<-c(2,3,4,5,6)
vector1
vector2<-c("a", "fancy", "vector")
vector2
vector3<-c(TRUE, FALSE, TRUE, FALSE, FALSE)
vector3
```

## Vector operations

```{r}
vector1
2 * vector1
vector3
vector3==FALSE
```

## Vector indexing

```{r}
vector2
vector2[1]
vector2[2]
vector2[3]
```

## Operations and vector indexing

```{r}
vector1
vector1[2]
vector1[2] + 3
```

## Functions!

R has loads and loads of functions. 

- Functions run a fixed set of operations on some argument(s)
- Functions return a value that can be assigned to an object
- Functions take the general form *function(arguments)*

## Functions!

```{r}
vector1
min(vector1)
max(vector1)
mean(vector1)
sum(vector1)
```

## Functions can work together

```{r}
sum(vector1)
length(vector1)
sum(vector1)/length(vector1)
mean(vector1)
```

## We can even write our own!

```{r}
redundantMean<-function(x){
  n<-length(x)
  sum_x<-sum(x)
  xbar<-sum_x/n
  return(xbar)
}
redundantMean(vector1)
```

# Questions?

## Data frames 

```{r}
### load the book package
library(qss)
### attach the packaged UNpop data
data(UNpop)
head(UNpop)
```

## As super-matrixes (matrices?)

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
UNpop[1,1]
UNpop[2,2]
```

## Other means of indexing

```{r}
UNpop[,1]
UNpop[1,]
UNpop$year
UNpop[,"world.pop"]
```
## Some convenient data.frame functions

```{r}
summary(UNpop)
```

Other useful functions: *nrow(), head(), tail(), str(), ncol(), View()*

## How to do operations over a data.frame column (the typical approach)

```{r}
mean(UNpop$world.pop)
UNpop$world.pop/UNpop$world.pop[1]
```

## We can also do this using tidyverse functions

```{r}
library(tidyverse)
UNpop %>% 
  mutate(pct_change = 
           world.pop/min(world.pop))
```

## Common tidyverse functions

- select() selects columns by name
- filter() filters rows by condition
- mutate() creates new columns based on arguments
- rename() renames columns
- arrange() reorders rows based on value

# We'll experiment with these in lab
