---
title: "1. Intro to Intro to Stats"
author: "Frank Edwards"
format:
  revealjs: 
    theme: dark
    embed-resources: true
execute: 
  echo: true
---

## Syllabus / schedule

<https://f-edwards.github.io/intro_stats>

## Books for class

Required: Gelman, Hill, and Vehtari. Regression and Other Stories (ROS). 2024. <https://avehtari.github.io/ROS-Examples/index.html>

Recommended: Alexander. Telling Stories with Data. 2023. <https://tellingstorieswithdata.com/>

## Navigating the schedule

<https://f-edwards.github.io/intro_stats>

## Software installation

- Install R <https://cran.r-project.org/>
- Install RStudio <https://posit.co/download/rstudio-desktop/>
- On windows Git Bash: <https://git-scm.com/install/windows>

## Open science basic principles

-   Science should be reproducible
-   Scientific decision-making should be transparent
-   Free and open source software are the best foundations for reproducible and open research

# RStudio walkthrough

# The script and the console

# Let's work with R

## Using the console

- Try some simple math operations with: + - / *  
- Now try some operations that use: ^ ()
- Try sqrt(); try log()

## Objects in R

For the following, please work in your RStudio Console

-   In R, we use the assignment operator `<-` to assign values to objects
-   Try to create an object named $x$ and assign the value 3 to the object
-   Confirm that $x$ is equal to three using the `==` operator
-   Compute $\frac{x+3}{2x^2}$

## Objects in R

- Create a new object named $y$ that is equal to $3x$
- Confirm that $y$ is equal to $3x$ using the `==` operator
- Print both y and 3x to the console

## The Environment

- List all objects in your environment with `ls()` 
- Compare the results to your 'Environment` tab in RStudio
- Next, let's restart our session of R and see what happens to our Environment 
- (Session Menu -\> Restart R or Shift + Cmd + 0)

## Reproducible research: the environment is a sandbox, treat it that way

- The environment should always be treated as temporary
- It disappears when you close R
- For now, you never need to save the environment
- You can use the environment for quick checks on objects or operations

## Reproducible research: the importance of scripts

- Scripts provide line-by-line instructions for everything you do in your analysis.
- ALWAYS DO YOUR WORK IN SCRIPTS!
- Scripts allow your entire analysis to be reproduced by simply running the program
- To make a new script: Cmd+Shift+N, or File -\> New File -\> R Script

## File paths and the command line

We need to choose a good place to save this cool new script we are making

-   Click the 'terminal' tab and let's check out how file paths work from the command line
-   Basic unix command line tools: cd; ls; mkdir
-   Navigate to your root directory with `cd ~`
-   Use `ls` to identify a good place for you to store your stats course work
-   create a new directory called `intro_stats` using mkdir

## Your class folder and path

- Now save this script as `lab1.R` in `./intro_stats/`
- Make note of its exact file path (the details before the cursor on your terminal)
- Create a new directory inside `intro_stats` called `hw`
- Use these directories all semester!

## RStudio projects

- Projects keep your scripts pointed to the right directories
- Create a new project for intro_stats now
- Always work in a project

## Back to our script

Let's repeat the operations we did before in the console, but now in the script
- On line 1: create an object named $x$ and assign the value 3 to the object
- On line 2: Confirm that $x$ is equal to three using the `==` operator
- On line 3: Create a new object named $y$ that is equal to $3x$
- On line 4: print the object y by simply calling `y`

## Sourcing and confirming output

-   Now run your program! Click 'source' or use cmd+shift+return to run the script
-   Check out your environment tab in RStudio, what do you see?
-   Restart R again: cmd+shift+0
-   Now source your script again. Reproducibility!

# Congrats you wrote a program!

## Let's document your program

- We use # to write code comments
- Before each line of code in your script, write a brief comment explaining what the line of code will do in plain english

## Packages in R
- Packages extend the functionality of R (dramatically!)
- Thousands of FOSS packages available through CRAN
- Try it:
  - On the console: `install.packages("tidyverse")
  - you only have to install once!

## Loading packages in R

- Use the `library("packagename")` format to import a package
- Try it with tidyverse
- Package loading will usually be the first few lines of your scripts

## Quarto and plaintext writing

Quarto is a powerful and flexible tool for writing reports, slides, and websites. It's very easy to use in RStudio

<https://quarto.org>

## Trying out quarto for a report

In RStudio

Click File-> New File -> Quarto Document

## Tour of quarto

- YAML header
- Markdown plain text
- Code chunks
- Rendering output

## LLMs and writing code

- You are here to learn how to do statistical analysis
- LLM use has been shown to impede learning and skill retention
- LLMs are very good at writing computer code

## LLM policy

1. You do not ask an LLM to complete an entire assignment or homework question.
2. You first try to solve a problem yourself. 
3. Only use LLMs for specific advice on technical problems
4. Clearly indicate where you used AI in your assignment with the flag AI support used
5. Include a full transcript of your LLM sessions. 

## Local LLMs

- https://ollama.com/download
- Intsall Gemma 4
- Configure Posit assistant

## LLM Usage

- Example session
- Export log file

# Introducing vectors and data.frames

## Vectors

## R works with objects 

```{r}
a<-2
a<-a+1
```

## R works with objects 

```{r}
a
```

## Objects can take many types

```{r}
a<-2
class(a)
```

## Objects can take many types

```{r}
b<-"howdy"
class(b)
```

## Objects can take many types

```{r}
c<-TRUE
class(c)
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector1<-c(2,3,4,5,6)
vector1
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector2<-c("a", "fancy", "vector")
vector2
```

## Vectors

```{r}
vector3<-c(TRUE, FALSE, TRUE, FALSE, FALSE)
vector3
```

## Vectorized operations

```{r}
vector1
2 * vector1
```

## Vectorized operations

```{r}
vector3
vector3==FALSE
```

## Vector indexing

```{r}
vector2
```

## Vector indexing

```{r}
vector2
vector2[1]
```

## Vector indexing

```{r}
vector2
vector2[2]
```

## Vector indexing

```{r}
vector2
vector2[3]
```

## Operations and vector indexing

```{r}
vector1
vector1[2] + 3
```

## Functions!

R has loads and loads of functions. 

- Functions run a fixed set of operations on some argument(s)
- Functions return a value that can be assigned to an object
- Functions take the general form *function(arguments)*

## Some common function in R for vectors

```{r}
vector1
max(vector1)
min(vector1)
```

## Some common function in R for vectors

```{r}
vector1
length(vector1)
```

## Some common function in R for vectors

```{r}
vector1
mean(vector1)
```

## Some common function in R for vectors

```{r}
vector1
sum(vector1)
```

## Functions can work together

```{r}
sum(vector1)
length(vector1)
# this is the same as mean()
sum(vector1)/length(vector1)
```

## We can even write our own!

```{r}
redundantMean<-function(x){
  n<-length(x)
  sum_x<-sum(x)
  xbar<-sum_x/n
  return(xbar)
}
redundantMean(vector1)
```

## Exercises

Create a new vector (any name is fine) that contains the elements 4, 6, 22, 3, 5

- Access the 3rd element of your vector
- Use the `length` function to see how long your vector is
- Use the `mean` function to compute the mean of your vector
- Use `sum` to compute sum; confirm that this is equal to mean * length

## Data frames 

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[1,1]
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[2,2]
```

## Other means of indexing

```{r}
iris[,1]
```

## Other means of indexing

```{r}
iris[1,]
```

## Other means of indexing

```{r}
iris$Petal.Length
```

## Other means of indexing

```{r}
iris$Petal.Length[3]
```

## Other means of indexing

```{r}
iris[1, "Petal.Length"]
```

## Some convenient data.frame functions

```{r}
summary(iris)
```

## Some convenient data.frame functions

```{r}
nrow(iris)
ncol(iris)
dim(iris)
```

## Some convenient data.frame functions

```{r}
head(iris)
```

## Some convenient data.frame functions

```{r}
tail(iris)
```

## Some convenient data.frame functions

```{r}
names(iris)
```

## How to do operations over a data.frame column

```{r}
mean(iris$Petal.Length)
sum(iris$Petal.Length)/nrow(iris)
```

## How to do operations over a data.frame column

```{r}
iris$Petal.Length/2
```

## Scatterplots

We make a *scatterplot* using `plot(var1, var2)`. 

```{r}
plot(iris$Petal.Length, iris$Petal.Width)
```


## Homework 1

<https://github.com/f-edwards/intro_stats/blob/master/hw/HW1.qmd>
