---
title: "3: Introduction to probability"
author: "Frank Edwards"
format:
  revealjs: 
    theme: dark
    embed-resources: true
    incremental: true
execute: 
  echo: true
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(pander)
library(qss)

options(xtable.comment = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "tiny")
```

## Probability

**How often, on average, does an event occur? **

- Probability is a set of tools for describing randomness. 

- Probability helps us sort signal (patterns) from noise.

## Two core theories

*Frequentist:* Probability is the proportion of times an event occurs if we repeat an experiment under the same conditions many times

## Consider a simple experiment

- What is the probability of obtaining a heads on a single fair coin flip?

. . . 

```{r size = "tiny"}
flip_n_coins<-function(x){
  flip<-rbinom(x, 1, 0.5)
  flip<-ifelse(flip==1, "Heads", "Tails")
  return(flip)
}
```

. . .

```{r}
flip_n_coins(1)
```

. . . 

```{r}
flip_n_coins(1)
```

. . . 

```{r}
flip_n_coins(1)
```

. . . 

What is $\frac{\sum_1^n x_i}{n}$ where $x_i=1$ when the coin is heads?

## Probability as the average outcome under repeat experiments

```{r}
sum((flip_n_coins(5)=="Heads")/5)
```

. . . 

```{r}
sum((flip_n_coins(20)=="Heads")/20)
```

. . . 

```{r}
sum((flip_n_coins(50)=="Heads")/50)
```

. . . 

```{r}
sum((flip_n_coins(1000)=="Heads")/1000)
```

. . . 

```{r}
sum((flip_n_coins(100000)=="Heads")/100000)
```

## Frequentist probability

If $n_x$ is the number of heads, and $n_t$ is the number of coin flips, then the probability of heads is

$$P(x) \approx \frac{n_x}{n_t}$$
$$P(x) = \lim_{n_t\to\infty} \frac{n_x}{n_t}$$

## Bayesian probability

*Bayesian:* Probability is a subjective judgment about the likelihood that an event occurs, with endpoints at 0 (never occurs) and 1 (always occurs). 

## Definitions

*Deterministic* processes do not include randomness. For example, if I drop a ball, it will fall (because gravity).

. . . 

*Stochastic* events include randomness. For example, if I flip a fair coin, it will be heads half of the time. 

. . . 


Nearly all social processes are have random components, and can be treated as *stochastic*. 

## Definitions and axioms

- Experiment: an action that produces stochastic events
- Sample space ($\Omega$): a set of all possible outcomes of the experiment 
- Event: a subset of the sample space

## Example: coin flips

- Experiment
    
    1. flip a coin
    2. roll a dice

. . . 

- Sample space $\Omega$

    1. {Heads, Tails}
    2. {1,2,3,4,5,6}

. . . 

- Event

    1. Heads, tails, not heads, heads or tails, heads and tails
    2. 3, even number, anything but 6

## Probability axioms

1. The probability of any event A is non-negative: $P(A)\geq0$
2. If an experiment is conducted, the probability that one of the outcomes in the sample space occurs is 1: $P(\Omega)=1$ 
3. Addition rule: If events A and B are mutually exclusive: $P(A \textsf{ or } B) = P(A) + P(B)$

## Probability that an event doesn't occur

$$1-P(\textsf{not } A) = P(A)$$

. . . 

If $P(\textsf{rolling a 6}) = \frac{1}{6}$ then $P(\textsf{not rolling a 6}) = 1 - \frac{1}{6}=\frac{5}{6}$

## Law of total probability

$$P(A) = P(A \textsf{ and } B) + P(A \textsf{ and not } B)$$

. . . 

Assume the sample space for A is [eats pizza, does not eat pizza], and the sample space for B is [happy, unhappy].

. . . 

If $P(\textsf{eats pizza}) = 0.5$ and $P(\textsf{eats pizza, happy}) = 0.4$, then $P(\textsf{eats pizza, unhappy})  = 0.1$

## General addition rule

$$P(A \textsf{ or } B) = P(A) + P(B) - P(A \textsf{ and } B)$$

. . . 

If $P(\textsf{happy})=0.5$, then $P(\textsf{eats pizza or happy})=0.6$

# Three kinds of probability

## Joint probability

The joint probability of two events (A and B) occurring is expressed as

$$P(A \textsf{ and } B)$$

## Marginal probability

The marginal probability of an event B is 

$$P(B)$$

## Conditional probability

The conditional probability of event $A$ occurring given that event $B$ occurred is 

$$P(A|B) = \frac{P(A \textsf{ and } B)}{P(B)}$$

# Working with some real data

## Voter files

```{r}
data("FLVoters")
voters<-na.omit(FLVoters)
head(voters)
```

## Marginal probability

What is the probability that a randomly sampled voter in the population is Black: $P(Black) = ?$

```{r}
voters |> 
  count(race) |> 
  mutate(p = n/sum(n))
```

. . . 

Is a woman: $P(Woman)= ?$

```{r}
voters |> 
  count(gender) |> 
  mutate(p = n/sum(n))
```

## Joint probability

What is the probability that a voter is a Black woman: $P(\textsf{Black and woman}) = ?$

```{r}
voters |> 
  count(gender, race) |> 
  mutate(n = n/sum(n)) |> 
  pivot_wider(names_from = gender, values_from = n)
```

## What is the probability that a voter is a woman? 

```{r echo = FALSE}
voters |> 
  count(gender, race) |> 
  mutate(n = n/sum(n)) |> 
  pivot_wider(names_from = gender, values_from = n)
```

Use the law of total probability:

$$P(A) = P(A \textsf{ and } B) + P(A \textsf{ and not } B)$$

. . . 

put differently, for all categories of B $i$:

$$P(A) = \sum_{i=1}^n{P(A \textsf{ and } B_i)}$$

## Conditional probability

If a voter is a man, what is the probability that he is Asian: $P(\textsf{Asian|man})= ?$

```{r}
voters |> 
  filter(gender=="m") |> 
  count(race) |> 
  mutate(p=n/sum(n))
```

## Conditioning on more than one variable

What is the probability that a male voter over age 60 is white?

$$P(\textsf{white|male and over 60)}$$

```{r}
voters |> 
  mutate(over60=age>60, white = race=="white") |> 
  count(over60, gender, white) |> 
  mutate(n=n/sum(n)) |> 
  pivot_wider(names_from = gender, values_from = n) 
```

## Independence

Two events are independent if knowledge of one event gives us no information about the other event. 

. . . 

$P(A|B)=P(A)$ and $P(B|A)=P(B)$

. . . 

$$ A \perp B $$ 

if and only if 

$$P(A \textsf{ and } B) = P(A)P(B)$$

## Bayes' rule

How should we change our beliefs after we make observations about the world? 

. . . 

$$P(A|B) = \frac{P(B|A)P(A)}{P(B)}$$

## Prior beliefs and evidence

If we have a *prior* belief that event A has $P(A)$ chance of occurring, we observe some data, $B$, we update our beliefs and obtain a *posterior probability* $P(A|B)$. 

$$P(A|B) = \frac{P(B|A)P(A)}{P(B)}$$

## Example: Detecting breast cancer

How good is a mammogram at detecting breast cancer?

. . . 

One percent of women have breast cancer. 80 percent of people who have cancer and take a mammogram test positive. 9.6 percent of mammograms are a false positive 

. . . 

If you take a mammogram and get a positive result, what is the probability that you have breast cancer?

## Rewriting as probabilities

One percent of women have breast cancer

$$P(\textsf{Cancer}) = 0.01$$ 

. . . 

80 percent of people who have cancer and take a mammogram test positive

$$P(\textsf{Test positive|Cancer}) = 0.8$$ 

. . . 

9.6 percent of people who take a mammogram get a positive result when they do not have breast cancer

$$P(\textsf{Test positive|No cancer}) = 0.096$$

## Using Bayes' rule

The prior probability of having cancer is 0.01. How should we update our belief that someone has cancer based on a positive test?

$$P(A|B) = \frac{P(B|A)P(A)}{P(B)}$$

. . . 

Using the law of total probability, we can rewrite the denominator as:

$$P(B) = P(B|A)P(A) + P(B|\textsf{ not A})P(\textsf{not }A)$$

## Using Bayes' rule

We can apply Bayes' rule for A = Cancer, B = positive test:

$$P(\textsf{Cancer|Test positive})=$$ 

$$\frac{P(\textsf{Test positive|Cancer})P(\textsf{Cancer})} {P(\textsf{Test positive)}}$$

$$P(\textsf{Cancer|Test positive})=\frac{0.8 \times 0.01}{0.8 \times 0.01 +  0.096 \times 0.99}$$

```{r}
(0.8 *  0.01)/(0.8 *  0.01 + 0.096 * 0.99)
```

# Random Variables

## Random Variables

- Random variables assign values to events 
- Each value is mutually exclusive 
- The set of all values is exhaustive (the sample space $\Omega$) 
- Discrete random variables take a finite number of values (e.g. TRUE, FALSE) 
- Continuous random variables are real numbers, and take on an infinite number of values 

## The simplest random variable: the binary Bernoulli

- Any random variable with two values is called a Bernoulli random variable. 

- Bernoulli (binary) variables are typically represented as $[0,1]$ or $[T,F]$.

## A coin flip as a Bernoulli random variable

A coin flip can be defined as a discrete random variable $X$ 

- If the coin lands on heads, $X=1$
- If the coin lands on tails, $X=0$ 
- The probability of a Bernoulli variable is the probability of success, or $X=1$ 
- $P(X=1) = p$

## Random variable (probability distribution) notation

$$X \sim Bernoulli(p)$$ 

. . . 

Reads: X is a Bernoulli distributed random variable with probability p 

## Probability mass

The probability mass function (PMF) of a variable $X$ is defined as the probability that a variable takes on a particular value $x$. 

$$PMF(x) = P(X=x)$$

## PMF for a coin flip

$$PMF(X=1)=p=0.5$$

$$PMF(X=0)=1-p=0.5$$

```{r echo=FALSE, fig.height=3}

temp_dat<-data.frame(x=c("0","1"), p=0.5)
ggplot(temp_dat, aes(x=x, y=p)) +
  geom_col()

```

## The probability mass function for passing the bar in NJ (p=0.653)

$$PMF(X=1)=p=0.653$$

$$PMF(X=0)=1-p=0.347$$

```{r echo=FALSE, fig.height=3}
temp_dat<-data.frame(x=c("0","1"), p=c(0.347,0.653))
ggplot(temp_dat, aes(x=x, y=p)) +
  geom_col()
```

## Describing a probability distribution: cumulative probability

- How likely is a variable to take a value less than or equal to a specified value? 

- We define the cumulative distribution function as the sum of all probabilities up to a value x

- $$CDF(X) = P(X \leq x) = \sum_{k \leq x}^x PMF(k)$$

- The CDF always ranges from 0 to 1, and never decreases as $x$ increases. 

## Uniform random variables

- Uniform random variables have an equal probability of taking any real value within a given interval $[a,b]$. 

- What does $X \sim Uniform(0,10)$ look like?

## Uniform random variables

Let's simulate it! 10 draws

```{r fig.height=3}
unif_draws<-runif(10, min=0, max=10)
unif_draws
hist(unif_draws, freq=F)
```

## Uniform random variable: 100 draws

```{r echo=F}
unif_draws<-runif(100, min=0, max=10)
hist(unif_draws, freq=F)
```

## Uniform random variable: 10000 draws

```{r echo=F}
unif_draws<-runif(10000, min=0, max=10)
hist(unif_draws, freq=F)
```

## PDF for $X \sim \textrm{Uniform}(0,10)$

```{r echo = F}
b<-10; a<-0
p<-1/(b-a)
temp_dat<-data.frame(x1=c(-5, 0, 10),
                     x2=c(0, 10, 15),
                     y1=c(0, p, 0),
                     y2=c(0, p, 0))
ggplot(temp_dat, aes(x=x1, xend=x2, y=y1, yend=y2)) + 
  geom_segment() + 
  xlab("x") + 
  ylab("Density of P(x=X)")
```

## CDF for $X \sim \textrm{Uniform}(0,10)$

```{r echo = F}
b<-10; a<-0
p<-1/(b-a)
temp_dat<-data.frame(x1=c(-5, 0, 10),
                     x2=c(0, 10, 15),
                     y1=c(0, 0, 1),
                     y2=c(0, 1, 1))
ggplot(temp_dat, aes(x=x1, xend=x2, y=y1, yend=y2)) + 
  geom_segment() + 
  xlab("x") + 
  ylab("P(x<=X)")
```

## The expectation of a random variable

- The expectation of a random variable $E(X)$ is the mean of a random variable. 


- For a discrete variable, the expectation is the sum of all values of $x$ weighted by their probability, given by the PDF $f(x)$.

- $$E(X) = \sum_x x \times f(x)$$

## Variance and standard deviation 

- The standard deviation $sd$ is

- $$sd = \sqrt{\frac{1}{n}\sum_{i=1}^n{(x_i-\bar{x}})^2}$$

- And the sample variance is $sd^2$ 

## The Normal Distribution

The Normal (Gaussian) distribution is continuous, and takes on values from $[-\infty, \infty]$. It has two parameters, the mean $\mu$ and standard deviation $\sigma$ (or variance $\sigma^2$).

- $\mu$ determines the location of the distribution
- $\sigma$ determines the spread of the distribution

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1), 
               name = "Mu=0, sd=1")
n2<-data.frame(x=sequence, y=dnorm(sequence, 2, 1), 
               name = "Mu=2, sd=1")
n3<-data.frame(x=sequence, y=dnorm(sequence, 0, 2), 
               name = "Mu=0, sd=2")


plot_dat<-bind_rows(n1, n2, n3)

ggplot(plot_dat,
       aes(x = x, y = y, color = name)) + 
  geom_line() + 
  ylab("Density of P(X=x)")
```

## The Normal CDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=pnorm(sequence, 0, 1), 
               name = "Mu=0, sd=1")
n2<-data.frame(x=sequence, y=pnorm(sequence, 2, 1), 
               name = "Mu=2, sd=1")
n3<-data.frame(x=sequence, y=pnorm(sequence, 0, 2), 
               name = "Mu=0, sd=2")


plot_dat<-bind_rows(n1, n2, n3)

ggplot(plot_dat,
       aes(x = x, y = y, color = name)) + 
  geom_line() + 
  ylab("Density of P(X=x)")
```

## Special features of Normal distributions: 

- The sum of many random variables from other distributions are often Normal
- For $X \sim N(\mu, \sigma^2)$, $Z=X+c$ is also Normal: $Z \sim (\mu +c, \sigma^2)$
- $Z=cX$ is distributed $Z\sim N(c\mu, (c\sigma)^2)$
- Z-scores of a Normal random variable are $N(0,1)$

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) 
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,0),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(pnorm(0,0,1))
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=0, y=0.5), color = "blue", size=2)
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,-1),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(round(pnorm(-1,0,1),3))
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=-1, y=0.159), color = "blue", size=2)
```

## Z-scores and area under the curve

To obtain a z-score, we subtract the mean and divide by the standard deviation:

$$\textsf{z-score} = \frac{X-\mu}{\sigma}$$

For a Normal variable, z-scores are distributed $z \sim N(0,1)$ 

What does a z-score of 0 indicate?  -1?  2?

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-1,1),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 1 SD = ",round(pnorm(1,0,1) - pnorm(-1,0,1),3)))
```

##

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-2,2),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 2 SD = ",round(pnorm(2,0,1) - pnorm(-2,0,1),3)))
```

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-3,3),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 3 SD = ",round(pnorm(3,0,1) - pnorm(-3,0,1),3)))
```

## Useful probability distribution functions

```{r}
### Normal(0,1) probability density function
dnorm(x = 0, mean = 0, sd = 1)
### Normal(0,1) cumulative distribution function
pnorm(q = 0, mean = 0, sd = 1)
### Random draw from a normal(0,1) distribution
rnorm(n = 1, mean = 0, sd = 1)
### CDF position for a given probability (quantile)
qnorm(p = 0.75, mean = 0, sd = 1)
### You can also use dbinom(), pbinom(), rbinom(), qbinom()
```

