---
title: "2. Introducing data analysis in R"
author: "Frank Edwards"
date: "9/2/2019"
output: binb::metropolis

---

```{r include = F}
options(xtable.comment = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "tiny")
```


# Problems?

# Introducing: R!

## R is a calculator

```{r}
5+3
```

## R is a calculator

```{r}
5*3
```

## R is a calculator

```{r}
5/3
```

## R is a calculator

```{r}
5^3
```

## R can make comparisons

```{r}
5>3
```

## R can make comparisons

```{r}
5<=3
```

## R can make comparisons

```{r}
5==3
```

## R can make comparisons

```{r}
"a"=="a"
```

## R works with objects 

```{r}
a<-2
```

## R works with objects 

```{r}
a<-2
a+1
```

## R works with objects 

```{r}
a+1
b<-a+2
```

## R works with objects 

```{r}
a<-2
a<-a+1
```

## R works with objects 

```{r}
a
```

## Objects can take many types

```{r}
a<-2
class(a)
```

## Objects can take many types

```{r}
b<-"howdy"
class(b)
```

## Objects can take many types

```{r}
c<-TRUE
class(c)
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector1<-c(2,3,4,5,6)
vector1
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector2<-c("a", "fancy", "vector")
vector2
```

## Vectors

```{r}
vector3<-c(TRUE, FALSE, TRUE, FALSE, FALSE)
vector3
```

## Vectorized operations

```{r}
vector1
2 * vector1
```

## Vectorized operations

```{r}
vector3
vector3==FALSE
```

## Vector indexing

```{r}
vector2
```

## Vector indexing

```{r}
vector2
vector2[1]
```

## Vector indexing

```{r}
vector2
vector2[2]
```

## Vector indexing

```{r}
vector2
vector2[3]
```

## Operations and vector indexing

```{r}
vector1
vector1[2] + 3
```

## Functions!

R has loads and loads of functions. 

- Functions run a fixed set of operations on some argument(s)
- Functions return a value that can be assigned to an object
- Functions take the general form *function(arguments)*


## Some common function in R for vectors

```{r}
vector1
max(vector1)
min(vector1)
```


## Some common function in R for vectors

```{r}
vector1
length(vector1)
```

## Some common function in R for vectors

```{r}
vector1
mean(vector1)
```

## Some common function in R for vectors

```{r}
vector1
sum(vector1)
```

## Functions can work together

```{r}
sum(vector1)
length(vector1)
sum(vector1)/length(vector1)
```

## We can even write our own!

```{r}
redundantMean<-function(x){
  n<-length(x)
  sum_x<-sum(x)
  xbar<-sum_x/n
  return(xbar)
}
redundantMean(vector1)
```

# Questions?

## Data frames 

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[1,1]
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[2,2]
```

## Other means of indexing

```{r}
iris[,1]
```

## Other means of indexing

```{r}
iris[1,]
```

## Other means of indexing

```{r}
iris$Petal.Length
```

## Other means of indexing

```{r}
iris$Petal.Length[3]
```

## Other means of indexing

```{r}
iris[1, "Petal.Length"]
```

## Some convenient data.frame functions

```{r}
summary(iris)
```

## Some convenient data.frame functions

```{r}
nrow(iris)
ncol(iris)
dim(iris)
```

## Some convenient data.frame functions

```{r}
head(iris)
```

## Some convenient data.frame functions

```{r}
tail(iris)
```

## Some convenient data.frame functions

```{r}
names(iris)
```

## How to do operations over a data.frame column

```{r}
mean(iris$Petal.Length)
sum(iris$Petal.Length)/nrow(iris)
```

## How to do operations over a data.frame column

```{r}
iris$Petal.Length/2
```

# Break

## Introducing Tidyverse

- Tidyverse is a collection of packages that work together to make `R` work a bit more like `sql`

- These features make routine data manipulation tasks FAR easier than they are in base `R`

## Your first line of code in nearly every script from now on

```{r}
library(tidyverse)
```


## The basics of data manipulation: filter()

`filter()` allow us to pick observations based on their values

```{r size = "tiny"}
filter(iris, Sepal.Length == 5.1)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Sepal.Length<4.5)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Sepal.Length>=7.7)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Species == "versicolor" , Sepal.Length<5.5)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Species != "setosa")
```


## The basics of data manipulation: arrange()

Arrange reorders rows

```{r size = "tiny"}
arrange(iris, Sepal.Width)
```

## arrange() can sort ascending and descending

```{r size = "tiny"}
arrange(iris, desc(Sepal.Width))
```

## The basics of data manipulation: select()

`select()` allows us to focus on a subset of columns

```{r size = "tiny"}
select(iris, Species, Petal.Length)
```

## select() can explicitly drop columns

```{r size = "tiny"}
select(iris, -Sepal.Length, -Sepal.Width)
```

## rename() renames columns

```{r size = "tiny"}
rename(iris, type = Species)
```

## Basics of data manipulation: mutate()

`mutate()` adds new columns, which can be functions of existing columns

```{r size = "tiny"}
mutate(iris, Petal.Size = Petal.Length + Petal.Width)
```

## mutate() can create many new variables at once

```{r size = "tiny"}
mutate(iris, 
       Petal.Size = Petal.Length + Petal.Width,
       Sepal.Size = Sepal.Length + Sepal.Width,
       Petal.Sepal.Ratio = Petal.Size/Sepal.Size)
```

## Piping commands together

Tidyverse uses a special symbol, called a pipe `%>%` to string together commands. Cmd + shift + m will make one.

## Piping commands

```{r size = "tiny"}
iris %>% 
  filter(Species == "versicolor") %>% 
  select(Sepal.Length, Petal.Length, Species) %>% 
  mutate(Sepal.Petal.Ratio = Sepal.Length / Petal.Length)
```

## Summarizing data

`summarize()` uses a variety of summary functions over the data

```{r}
iris %>% 
  summarize(mean.pl = mean(Petal.Length),
            min.pl = min(Petal.Length),
            max.pl = max(Petal.Length))
```

## But summarize() is more powerful with its buddy, group_by()

`group_by()` groups the data by individual groups within the data

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(mean.pl = mean(Petal.Length),
            min.pl = min(Petal.Length),
            max.pl = max(Petal.Length))
```


# Output in R

## Writing to files

The most straightforward way to get output from R is to write to a file

```{r size = "tiny"}
iris_summary<-iris %>% 
  group_by(Species) %>% 
  summarize(mean.pl = mean(Petal.Length),
            min.pl = min(Petal.Length),
            max.pl = max(Petal.Length))

write_csv(iris_summary, "iris_summary.csv")
```

## RMarkdown

- RMarkdown allows us to combine code and text in one document (no copying and pasting needed!)
- With the proper workflow, you can do all your academic writing in one place.
- Let's work through some demos.

## HW 2

- Read chapter 5 of R for Data Science, Wickham (https://r4ds.had.co.nz/transform.html)
- Complete the following exercises: 5.2.4 (Question 1); 5.3.1 (all); 5.4.1 (Questions 1 and 2); 5.5.2 (Questions 1-5); 5.6.7 (Question 5); 5.7.1 (Questions 1, 2, 3, 4, 7)

