---
title: "4: Normal distributions, correlations"
author: "Frank Edwards"
format:
  revealjs: 
    theme: dark
    embed-resources: true
    incremental: true
execute: 
  echo: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(pander)
```

# Random Variables

## Random Variables

- Random variables assign values to events 
- Each value is mutually exclusive 
- The set of all values is exhaustive (the sample space $\Omega$) 
- Discrete random variables take a finite number of values (e.g. TRUE, FALSE) 
- Continuous random variables are real numbers, and take on an infinite number of values 

## The expectation of a random variable

- The expectation of a random variable $E(X)$ is the mean. 

## Variance and standard deviation 

- The standard deviation $sd$ measures the average difference of observations from the mean

. . . 

$$sd = \sqrt{\frac{1}{n}\sum_{i=1}^n{(x_i-\bar{x}})^2}$$

- Sometimes we use the variance, which is $sd^2$ 

## The Normal Distribution

The Normal (Gaussian) distribution is continuous, and takes on values from $[-\infty, \infty]$. It has two parameters, the mean $\mu$ and standard deviation $\sigma$ (or variance $\sigma^2$).

- $\mu$ determines the location of the distribution
- $\sigma$ determines the spread of the distribution

# The location parameter

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=0, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0.5, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=0.5, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 1, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=1, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 1.5, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=1.5, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 2, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=2, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 2.5, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=2.5, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 3, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=3, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 3.5, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=3.5, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 4, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=4, sigma = 1")
```

# The variance parameter

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)") + 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4))+ 
  labs(subtitle="mu=0, sigma = 1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.1))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.1")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.2))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.2")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.3))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.3")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.4))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.4")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.5))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.5")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.6))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.6")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.7))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.7")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.8))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.8")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1.9))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 1.9")
```

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 2))

plot_dat<-bind_rows(n1)

ggplot(plot_dat,
       aes(x = x, y = y)) + 
  geom_line() + 
  ylab("Density of P(X=x)")+ 
  coord_cartesian(xlim = c(-7, 7), ylim = c(0, 0.4)) + 
  labs(subtitle="mu=0, sigma = 2")
```


## The Normal Cumulative Distribution Function 

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=pnorm) 
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -3
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -2.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -2
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -1.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -1
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -0.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- -0
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 0.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 1
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 1.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 2
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 2.5
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Interpreting the PDF and CDF

```{r echo = F}
s <- 3
p1 <- ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,s),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(c(paste(s, "sigma,", "p=",round(pnorm(s,0,1),3)))) + 
  labs(subtitle = "Probability Density Function")

p2 <- ggplot(data.frame(x=c(-4,4), s = s, y = pnorm(s, 0, 1)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=s, y=y), color = "blue", size=2) + 
  labs(subtitle = "Cumulative Distribution Function")

gridExtra::grid.arrange(p1, p2)
```

## Z-scores and area under the curve

- To obtain a z-score, we subtract the mean and divide by the standard deviation:

. . . 

$$\textsf{z-score} = \frac{X-\mu}{\sigma}$$

- For a Normal variable, z-scores are distributed $z \sim N(0,1)$ 

- What does a z-score of 0 indicate?  -1?  2?

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-1,1),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 1 SD = ",round(pnorm(1,0,1) - pnorm(-1,0,1),3)))
```

##

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-2,2),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 2 SD = ",round(pnorm(2,0,1) - pnorm(-2,0,1),3)))
```

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-3,3),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 3 SD = ",round(pnorm(3,0,1) - pnorm(-3,0,1),3)))
```

## Special features of Normal distributions: 

- The sum of many random variables from other distributions are often Normal
- If X is a Normal random variable, then $Y=X+c$ is also Normal
  - $Y \sim (\mu +c, \sigma^2)$
- It works for multiplication too, as a scaling factor
  - $Y=cX$ is distributed $Y\sim N(c\mu, (c\sigma)^2)$
- Z-scores of a Normal random variable are $N(0,1)$

# Correlation

## Correlation defined

- Correlation measures the linear relationship between two variables $x$ and $y$

. . . 

$$cor_{xy} = \frac{\textrm{E}[(x-\bar{x})(y-\bar{y})]}{\sigma_x \cdot \sigma_y}$$

## Correlation and Z-scores

- It's helpful to have measures on the same scale

- A $z$-score measures the number of standard deviations an observation is away from the mean of a variable.

- $z\textrm{ score of }x_i = \frac{x_i-\bar{x}}{S_x}$

- In R, we can transform a numeric into a $z$-score using `scale()`

## Z-scores in R

```{r}
iris  |>  
  mutate(Sepal.Length.sc = scale(Sepal.Length)) |> 
  select(Sepal.Length, Sepal.Length.sc)
```

## Z-score transformed distributions have the same shape as the original data

```{r echo = F, message = F}
library(patchwork)
p1<-ggplot(iris,
           aes(x = Sepal.Length)) + 
  geom_histogram() + 
  labs(subtitle = "Untransformed")
p2<-ggplot(iris,
           aes(x = scale(Sepal.Length))) + 
  geom_histogram() + 
  labs(subtitle = "Z-transformed")

p1+p2
```

## Correlation

- Correlation measures the linear association between two variables. We often use the letter $r$ to denote a correlation.

. . . 

$$r(x,y) = \frac{1}{n}\sum_{i=1}^{n} \frac{x_i-\bar{x}}{S_x} \times \frac{y_i-\bar{y}}{S_y}$$

. . . 

$$ = E[z(x) \times z(y)] $$

. . . 

In R, you can use `cor()`

## Evaluate correlations using cor()

- Compute the correlation between Sepal.Length and Sepal.Width. What does it mean?
- Compute the correlation between Petal.Length and Petal.Width. What does it mean?
- Compute the correlation between Petal.Length and Sepal.Width. What does it mean?

# Interpreting relationships with scatterplots

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=0, to =1, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  facet_wrap(~cor)
```

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=0, to =1, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F) + 
  facet_wrap(~cor)
```

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=-1, to =0, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  facet_wrap(~cor)
```

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=-1, to =0, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F) + 
  facet_wrap(~cor)
```

## Return to iris

- Scatterplot Sepal.Width and Petal.Width
- How would you describe the relationship between the variables?
- Estimate the correlation
- Describe what you find using both the estimated correlation and your interpretation of the scatterplot

# Why we visualize

## Exercise

- load `data("anscombe")
- Compute correlations for each pair x1, y1; x2, y2; ...
- What do you find?
- Now scatterplot each pair using ggplot

## Homework prep

- Download the Uniform Crime Reports Arrests by Age, Sex, and Race 2022 data from NACJD in delimited format: https://www.icpsr.umich.edu/web/NACJD/studies/39063
- Move the data into a subdirectory of your intro_stats folder called ./data/

