---
title: "4. Vectors and intro to the Tidyverse"
author: "Frank Edwards"
output: binb::metropolis

---

```{r include = F}
options(xtable.comment = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "tiny")
```

## R works with objects 

```{r}
a<-2
a<-a+1
```

## R works with objects 

```{r}
a
```

## Objects can take many types

```{r}
a<-2
class(a)
```

## Objects can take many types

```{r}
b<-"howdy"
class(b)
```

## Objects can take many types

```{r}
c<-TRUE
class(c)
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector1<-c(2,3,4,5,6)
vector1
```

## Vectors

Vectors are one-dimensional arrays of values of any class

```{r}
vector2<-c("a", "fancy", "vector")
vector2
```

## Vectors

```{r}
vector3<-c(TRUE, FALSE, TRUE, FALSE, FALSE)
vector3
```

## Vectorized operations

```{r}
vector1
2 * vector1
```

## Vectorized operations

```{r}
vector3
vector3==FALSE
```

## Vector indexing

```{r}
vector2
```

## Vector indexing

```{r}
vector2
vector2[1]
```

## Vector indexing

```{r}
vector2
vector2[2]
```

## Vector indexing

```{r}
vector2
vector2[3]
```

## Operations and vector indexing

```{r}
vector1
vector1[2] + 3
```

## Functions!

R has loads and loads of functions. 

- Functions run a fixed set of operations on some argument(s)
- Functions return a value that can be assigned to an object
- Functions take the general form *function(arguments)*

## Some common function in R for vectors

```{r}
vector1
max(vector1)
min(vector1)
```

## Some common function in R for vectors

```{r}
vector1
length(vector1)
```

## Some common function in R for vectors

```{r}
vector1
mean(vector1)
```

## Some common function in R for vectors

```{r}
vector1
sum(vector1)
```

## Functions can work together

```{r}
sum(vector1)
length(vector1)
# this is the same as mean()
sum(vector1)/length(vector1)
```

## We can even write our own!

```{r}
redundantMean<-function(x){
  n<-length(x)
  sum_x<-sum(x)
  xbar<-sum_x/n
  return(xbar)
}
redundantMean(vector1)
```

## Exercises

Create a new vector (any name is fine) that contains the elements 4, 6, 22, 3, 5

- Access the 3rd element of your vector
- Use the `length` function to see how long your vector is
- Use the `mean` function to compute the mean of your vector
- Use `sum` to compute sum; confirm that this is equal to mean * length

## Data frames 

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
head(iris)
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[1,1]
```

## As super-matrices?

Recall that we can obtain any element $x_{ij}$ from a matrix $X$ with row index $i$ and column index $j$

```{r}
iris[2,2]
```

## Other means of indexing

```{r}
iris[,1]
```

## Other means of indexing

```{r}
iris[1,]
```

## Other means of indexing

```{r}
iris$Petal.Length
```

## Other means of indexing

```{r}
iris$Petal.Length[3]
```

## Other means of indexing

```{r}
iris[1, "Petal.Length"]
```

## Some convenient data.frame functions

```{r}
summary(iris)
```

## Some convenient data.frame functions

```{r}
nrow(iris)
ncol(iris)
dim(iris)
```

## Some convenient data.frame functions

```{r}
head(iris)
```

## Some convenient data.frame functions

```{r}
tail(iris)
```

## Some convenient data.frame functions

```{r}
names(iris)
```

## How to do operations over a data.frame column

```{r}
mean(iris$Petal.Length)
sum(iris$Petal.Length)/nrow(iris)
```

## How to do operations over a data.frame column

```{r}
iris$Petal.Length/2
```

## Exercises

Using `iris`

- Compute the sum of all petal lengths
- Compute the mean sepal width
- How many rows are in the data frame? 
- How many columns? 

## Introducing Tidyverse

- Tidyverse is a collection of packages that work together to make `R` work a bit more like `sql`

- These features make routine data manipulation tasks FAR easier than they are in base `R`

## Installing packages

On your console, run this command:

`install.packages("tidyverse")`

You only have to install packages once. 

## Your first line of code in nearly every script from now on

```{r}
library(tidyverse)
```

## The basics of data manipulation: filter()

`filter()` allow us to pick observations based on their values

```{r size = "tiny"}
filter(iris, Sepal.Length == 5.1)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Sepal.Length<4.5)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Sepal.Length>=7.7)
```

## filter() can take more than one argument

```{r size = "tiny"}
filter(iris, Species == "versicolor" , Sepal.Length<5.5)
```

## filter() can take many kinds of arguments

```{r size = "tiny"}
filter(iris, Species != "setosa")
```

## The basics of data manipulation: arrange()

Arrange reorders rows

```{r size = "tiny"}
arrange(iris, Sepal.Width)
```

## arrange() can sort ascending and descending

```{r size = "tiny"}
arrange(iris, desc(Sepal.Width))
```

## The basics of data manipulation: select()

`select()` allows us to focus on a subset of columns

```{r size = "tiny"}
select(iris, Species, Petal.Length)
```

## select() can explicitly drop columns

```{r size = "tiny"}
select(iris, -Sepal.Length, -Sepal.Width)
```

## rename() renames columns

```{r size = "tiny"}
rename(iris, type = Species)
```

## Basics of data manipulation: mutate()

`mutate()` adds new columns, which can be functions of existing columns

```{r size = "tiny"}
mutate(iris, Petal.Size = Petal.Length + Petal.Width)
```

## mutate() can create many new variables at once

```{r size = "tiny"}
mutate(iris, 
       Petal.Size = Petal.Length + Petal.Width,
       Sepal.Size = Sepal.Length + Sepal.Width,
       Petal.Sepal.Ratio = Petal.Size/Sepal.Size)
```

## Piping commands together

Tidyverse uses a special symbol, called a pipe `%>%` to string together commands. Cmd + shift + m will make one.

## Piping commands

The pipe `%>%` tells R to use `iris` for all commands that follow

```{r size = "tiny"}
iris %>% 
  filter(Species == "versicolor") 
```

## Piping commands

```{r size = "tiny"}
iris %>% 
  mutate(sepal2 = Sepal.Length + 2)
```

## Exercises

Using `iris` and the pipe operator `%>%`

- Use select to return only the petal variables and flower species
- Use filter to return only versicolor observations
- Use arrange to sort the data.frame by Sepal.Length
- Use mutate to compute the sum of petal lengths and widths
- Identify and return a data.frame containing only observations with the smallest observed petal width

## Piping many commands

```{r size = "tiny"}
iris %>% 
  filter(Species == "versicolor") %>% 
  select(Sepal.Length, Petal.Length, Species) %>% 
  mutate(Sepal.Petal.Ratio = Sepal.Length / Petal.Length)
```

## Summarizing data

`summarize()` uses a variety of summary functions over the data (e.g. `mean`, `min`, `max`, `sd`, etc.)

```{r}
iris %>% 
  summarize(mean.pl = mean(Petal.Length))
```

## But summarize() is more powerful with its buddy, group_by()

`group_by()` groups the data by individual groups within the data

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(mean.pl = mean(Petal.Length))
```

## We can use it to count too

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(n_obs = n())
```

## We can use it to count too

```{r}
iris %>% 
  filter(Sepal.Width<3) %>% 
  group_by(Species) %>% 
  summarize(n_obs = n())
```

## Exercises

- Which species has the largest average sepal length?
- Which species has the smallest average petal length?
- What is the average sepal width across all species?
- Which species has the highest number of observations with petal lengths less than the average petal length across all species?


