---
title: "5. More tidyverse and intro to ggplot"
author: "Frank Edwards"
output: binb::metropolis

---

```{r include = F}
options(xtable.comment = FALSE)
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = TRUE, size = "tiny")
library(tidyverse)
```

## Piping commands together

Tidyverse uses a special symbol, called a pipe `%>%` to string together commands. Cmd + shift + m will make one.

## Piping commands

The pipe `%>%` tells R to use `iris` for all commands that follow

```{r size = "tiny"}
iris %>% 
  filter(Species == "versicolor") 
```

## Piping commands

```{r size = "tiny"}
iris %>% 
  mutate(sepal2 = Sepal.Length + 2)
```

## Exercises

Using `iris` and the pipe operator `%>%`

- Use select to return only the petal variables and flower species
- Use filter to return only versicolor observations
- Use arrange to sort the data.frame by Sepal.Length
- Use mutate to compute the sum of petal lengths and widths
- Identify and return a data.frame containing only observations with the smallest observed petal width

## Piping many commands

```{r size = "tiny"}
iris %>% 
  filter(Species == "versicolor") %>% 
  select(Sepal.Length, Petal.Length, Species) %>% 
  mutate(Sepal.Petal.Ratio = Sepal.Length / Petal.Length)
```

## Summarizing data

`summarize()` uses a variety of summary functions over the data (e.g. `mean`, `min`, `max`, `sd`, etc.)

```{r}
iris %>% 
  summarize(mean.pl = mean(Petal.Length))
```

## But summarize() is more powerful with its buddy, group_by()

`group_by()` groups the data by individual groups within the data

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(mean.pl = mean(Petal.Length))
```

## We can use it to count too

```{r}
iris %>% 
  group_by(Species) %>% 
  summarize(n_obs = n())
```

## We can use it to count too

```{r}
iris %>% 
  filter(Sepal.Width<3) %>% 
  group_by(Species) %>% 
  summarize(n_obs = n())
```

## Exercises

- Which species has the largest average sepal length?
- Which species has the smallest average petal length?
- What is the average sepal width across all species?
- Which species has the highest number of observations with petal lengths less than the average petal length across all species?


# Visualizing the distribution of single variables (univariate visuals)

# Visuals for categorical variables

## Barplots

Show the count of rows in each value of a category

```{r fig.height = 3.5}
ggplot(mpg, 
       aes(y = class)) + 
  geom_bar()
```

## The anatomy of a basic ggplot() call

```{r eval = FALSE, size = "scriptsize"}
ggplot(mpg, ## object, usually a data.frame 
       aes(y = class)) + ## aesthetic variables, generally x, y, color, etc
  geom_bar() ## a geom to plot the aesthetics

## Note that + in ggplot() works the same way as %>% in tidyverse:
## It strings together commands, evaluated in sequence
```

## Practice

Using the `mpg` data.frame, make a barplot of `manufacturer`. Try changing the aesthetic from a `y` to an `x`. What happens?

Recall the anatomy of a ggplot call

```{r eval = F}
ggplot(DATA.FRAME.NAME,
       aes(AESTHETIC PARAMETERS)) + 
  GEOM
```

# Visuals for continuous variables

## Histograms

Histograms show the density of cases that fall within a given range
```{r fig.height = 3.5}
ggplot(mpg,
       aes(x = hwy)) + 
  geom_histogram()
```

## Histograms in base R

```{r}
hist(mpg$hwy)
```

# Practice

1. Make a histogram of city miles per gallon in the `mpg` dataset
2. What does this histogram tell us?

## Density plots

Densities are smoothed continuous histograms 
```{r fig.height = 3.5}
ggplot(mpg,
       aes(x = hwy)) + 
  geom_density()
```

## Another set of histograms/densities

```{r fig.height = 2}
ggplot(mpg, 
       aes(x = hwy)) + 
  geom_histogram()
```

```{r fig.height = 2}
ggplot(mpg, 
       aes(x = hwy)) + 
  geom_density()
```

# Practice

1. Create a histogram of `displ`
2. Create a density plot of `displ`
3. Which do you think is a more effective data visual? Why?
4. Create a histogram of city mpg
5. Create a histogram of highway mpg
6. Describe differences in these distributions.
7. Using a barplot, identify which vehicle class is most common in the data
8. Using a barplot, identify which manufacturer is least common in the data

