---
title: "Probability, 3"
author: "Frank Edwards"
output: binb::metropolis
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(pander)
library(qss)

options(xtable.comment = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "tiny")
```

## The expectation of a random variable

The expectation of a random variable $E(X)$ is the mean of a random variable. 

Be careful not to confuse $E(X)$ and $\bar{x}$. \pause

For a discrete variable, the expectation is the sum of all values of $x$ weighted by their probability, given by the PDF $f(x)$.

\[E(X) = \sum_x x \times f(x)\]

## Variance and standard deviation of a random variable

The standard deviation $sd$ is

\[sd = \sqrt{\frac{1}{n}\sum_{i=1}^n{(x_i-\bar{x}})^2}\]

And the sample variance is $sd^2$ \pause

## Variance defined

For a random variable $X$, the variance is defined via the expectation instead of sample mean

\[V(X) = E[\{X-E(X)\}^2]\] \pause

Note the similarities in the two equations

\[V(X) = \frac{1}{n}\sum_{i=1}^n{(x_i-\bar{x}})^2\]


## The Normal Distribution

The Normal (Gaussian) distribution is continuous, and takes on values from $[-\infty, \infty]$. It has two parameters, the mean $\mu$ and standard deviation $\sigma$ (or variance $\sigma^2$).

- $\mu$ determines the location of the distribution
- $\sigma$ determines the spread of the distribution

## The Normal PDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=dnorm(sequence, 0, 1), 
               name = "Mu=0, sd=1")
n2<-data.frame(x=sequence, y=dnorm(sequence, 2, 1), 
               name = "Mu=2, sd=1")
n3<-data.frame(x=sequence, y=dnorm(sequence, 0, 2), 
               name = "Mu=0, sd=2")


plot_dat<-bind_rows(n1, n2, n3)

ggplot(plot_dat,
       aes(x = x, y = y, color = name)) + 
  geom_line() + 
  ylab("Density of P(X=x)")
```

## The Normal CDF

```{r echo = F}
sequence<-seq(-8,8,0.001)
n1<-data.frame(x=sequence, y=pnorm(sequence, 0, 1), 
               name = "Mu=0, sd=1")
n2<-data.frame(x=sequence, y=pnorm(sequence, 2, 1), 
               name = "Mu=2, sd=1")
n3<-data.frame(x=sequence, y=pnorm(sequence, 0, 2), 
               name = "Mu=0, sd=2")


plot_dat<-bind_rows(n1, n2, n3)

ggplot(plot_dat,
       aes(x = x, y = y, color = name)) + 
  geom_line() + 
  ylab("Density of P(X=x)")
```

## Special features of Normal distributions: 

- The sum of many random variables from other distributions are often Normal
- For $X \sim N(\mu, \sigma^2)$, $Z=X+c$ is also Normal: $Z \sim (\mu +c, \sigma^2)$
- $Z=cX$ is distributed $Z\sim N(c\mu, (c\sigma)^2)$
- Z-scores of a Normal random variable are $N(0,1)$

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) 
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,0),
                geom="area",
                fill="blue",
                alpha = 0.5)+
  ggtitle(pnorm(0,0,1))
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=0, y=0.5), color = "blue", size=2)
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-4,-1),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(round(pnorm(-1,0,1),3))
```

## Area under the curve: interpreting the PDF and CDF

```{r echo = F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=pnorm) +
  geom_point(aes(x=-1, y=0.159), color = "blue", size=2)
```

## Z-scores and area under the curve

To obtain a z-score, we subtract the mean and divide by the standard deviation:

\[\textsf{z-score} = \frac{X-\mu}{\sigma}\]

For a Normal variable, z-scores are distributed $z \sim N(0,1)$ \pause

What does a z-score of 0 indicate? \pause -1? \pause 2?

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-1,1),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 1 SD = ",round(pnorm(1,0,1) - pnorm(-1,0,1),3)))
```

##

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-2,2),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 2 SD = ",round(pnorm(2,0,1) - pnorm(-2,0,1),3)))
```

## 

```{r echo=F}
ggplot(data.frame(x=c(-4,4)), aes(x)) +
  stat_function(fun=dnorm) +
  stat_function(fun=dnorm,
                xlim=c(-3,3),
                geom="area",
                fill="blue",
                alpha = 0.5) +
  ggtitle(paste("Mean +/- 3 SD = ",round(pnorm(3,0,1) - pnorm(-3,0,1),3)))
```

## Useful probability distribution functions

These will be useful for the homework!

```{r}
### Normal(0,1) probability density function
dnorm(x = 0, mean = 0, sd = 1)
### Normal(0,1) cumulative distribution function
pnorm(q = 0, mean = 0, sd = 1)
### Random draw from a normal(0,1) distribution
rnorm(n = 1, mean = 0, sd = 1)
### CDF position for a given probability (quantile)
qnorm(p = 0.75, mean = 0, sd = 1)
```

# RMarkdown basics
