---
title: "Correlation"
author: "Frank Edwards"
date: "10/26/2021"
output: binb::metropolis
---

```{r setup, include=FALSE}
rm(list=ls())
library(tidyverse)
library(MASS)
select<-dplyr::select
set.seed(1)

options(xtable.comment = FALSE)
theme_set(theme_bw())
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
knitr::opts_chunk$set(warning=FALSE, message=FALSE, tidy = F, size = "small")
```


## Correlation (math time): Z-scores

First, we need the variables to be comparable, so we transform them to be on a standard deviation scale.

A $z$-score scales a variable measures the number of standard deviations an observation is away from it's mean.

\[z\textrm{ score of }x_i = \frac{x_i-\bar{x}}{S_x} \]

Where $\bar{x}$ is the mean, and $S_x$ is the standard deviation of variable $x$. Z scores have a mean zero, and a range defined by the range of the data on a standard deviation scale. 

For a normally (Gaussian) distributed variable, this will typically range between $[-3,3]$ 

In R, we can transform a numeric into a $z$-score using \texttt{scale()}

## Z-scores in R

```{r}
iris %>% 
  mutate(Sepal.Length.sc = scale(Sepal.Length)) %>% 
  select(Sepal.Length, Sepal.Length.sc)
```

## Z-score transformed distributions have the same shape as the original data

```{r echo = F, message = F}
library(patchwork)
p1<-ggplot(iris,
           aes(x = Sepal.Length)) + 
  geom_histogram() + 
  labs(subtitle = "Untransformed")
p2<-ggplot(iris,
           aes(x = scale(Sepal.Length))) + 
  geom_histogram() + 
  labs(subtitle = "Z-transformed")

p1+p2
```

## Correlation

Correlation measures the degree to which two variables are associated with each other. We often use the letter $r$ to denote a correlation.

\[r(x,y) = \frac{1}{n}\sum_{i=1}^{n} \frac{x_i-\bar{x}}{S_x} \times \frac{y_i-\bar{y}}{S_y}\]
\[ = E[z(x) \times z(y)] \]

In R, you can use \texttt{cor()}

## Evaluate correlations using cor()

- Compute the correlation between Sepal.Length and Sepal.Width. What does it mean?
- Compute the correlation between Petal.Length and Petal.Width. What does it mean?
- Compute the correlation between Petal.Length and Sepal.Width. What does it mean?

# Bivariate visuals for continuous data: Scatterplots

## Make a scatterplot

```{r fig.height = 4}
ggplot(iris,
       aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point()
```

## Scatterplot ingredients

- a `data.frame` with two continuous variables
- `aes()` with an x and y parameter
- `geom_point()`

## Practice

- Scatterplot petal length on the x and petal width on the y
- Flip the axes (move length to y, width to x)

## Scatterploting with clusters

We often have bivariate measures that are *clustered*, or have some structure caused by a third (often categorical) variable. \pause

Do we see structure here? What could be causing it?

```{r size = "tiny", fig.height = 4}
ggplot(iris,
       aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point()
```

## Two solutions to plotting clustered data: aesthetics

Use the clustering variable to add another aesthetic element to our plot, like color

```{r size = "tiny", fig.height = 4}
ggplot(iris,
       aes(x = Sepal.Length, y = Sepal.Width,
           color = Species)) + 
  geom_point()
```

## Two solutions to plotting clustered data: facets

```{r size = "tiny", fig.height = 4}
ggplot(iris,
       aes(x = Sepal.Length, y = Sepal.Width)) + 
  geom_point() + 
  facet_wrap(~Species)
```

# Interpreting relationships with scatterplots

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=0, to =1, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  facet_wrap(~cor)
```

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=0, to =1, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F) + 
  facet_wrap(~cor)
```

## Correlation and scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=-1, to =0, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  facet_wrap(~cor)
```

## Correlation and Scatterplots

```{r echo = FALSE}
library(MASS)
select<-dplyr::select
cors<-seq(from=-1, to =0, by = 0.1)
cor_plot<-list()
for(i in 1:length(cors)){
  cor_plot[[i]] <- mvrnorm(50, 
                      mu = c(0,0), 
                      Sigma = matrix(c(1,cors[i],cors[i],1), ncol = 2),
                      empirical = TRUE)
  cor_plot[[i]]<-as.data.frame(cor_plot[[i]])
  cor_plot[[i]]$cor<-cors[i]
}
cor_plot<-bind_rows(cor_plot)

ggplot(cor_plot, aes(x=V1, y = V2)) + 
  geom_point() +
  geom_smooth(method = "lm", se = F) + 
  facet_wrap(~cor)
```

## Return to iris

- Scatterplot Sepal.Width and Petal.Width
- How would you describe the relationship between the variables?
- Estimate the correlation
- Describe what you find using both the estimated correlation and your interpretation of the scatterplot

# Why we visualize

## Exercise

- load `data("anscombe")
- Compute correlations for each pair x1, y1; x2, y2; ...
- What do you find?
- Now scatterplot each pair using ggplot

## Homework prep

- Download the Uniform Crime Reports Arrests by Age, Sex, and Race 2022 data from NACJD in delimited format: https://www.icpsr.umich.edu/web/NACJD/studies/39063
- Move the data into a subdirectory of your intro_stats folder called ./data/

