---
title: "Math camp problem set 1"
author: "Frank Edwards"
date: "8/30/2019"
output: pdf_document
---

Solve all problems by hand, and using R code. Simplify where possible using rules of exponents and logarithms.

\bigskip
\bigskip

1. $\sum_{k=1}^4(k-1)^2$

```{r}
## THIS IS A COMMENT
#### SOLUTION FOR QUESTION 1

### define k
k<-1:4
### sum with the function (k-1)^2
sum((k-1)^2)
```


\bigskip
\bigskip
\bigskip
\bigskip

2. $\prod_{i=2}^{10}\frac{i+1}{i}$

```{r}
i<-2:10
prod((i+1)/i)
```


\bigskip
\bigskip
\bigskip
\bigskip

3. $\log(e^2)$

```{r}
log(exp(2))
```


\bigskip
\bigskip
\bigskip
\bigskip

4. $e^4e^{10}$

```{r}
exp(4) * exp(10)
exp(4 + 10)
```


\bigskip
\bigskip
\bigskip
\bigskip

5. $10^3 10^{-2}$

```{r}
10^3 * 10^(-2)
10^(1)
10^3 * (1/10^2)
```

\bigskip
\bigskip
\bigskip
\bigskip

6. $400^{1/2}$

```{r}
400^(1/2)
sqrt(400)
```


\bigskip
\bigskip
\bigskip
\bigskip

7. $f(x) = 2x^2 + 3x - 7$. Solve for $x = 12$. For $x = -2$

```{r}
whateveriwant<-function(x){
  2*x^2 + 3*x -7
}
whateveriwant(12)
whateveriwant(-2)

2*12^2 + 3*12 -7
2*(-2)^2 + 3*(-2) -7 
```


\bigskip
\bigskip
\bigskip
\bigskip

8. Find the slope and y-intercept of the line that passes through (1,7) and (4,2) 

```{r}
x1<-1
x2<-4
y1<-7
y2<-2
m<-(y2-y1)/(x2-x1)
b<-y1-m*x1
m
b
```

