---
title: "HW1"
author: "YOUR NAME"
date: "`r Sys.Date()`"
format: html
---

# Practice with R

Load the tidyverse package

```{r}
library(tidyverse)
```

Load the iris data into your environment 

```{r}
data(iris)
```

1. Use the `mean` function to compute the average petal width in the sample

```{r}
# YOUR CODE GOES HERE
```

Your explanation goes here!

2. Use the `min` function to compute the smallest petal width in the data

```{r}
# YOUR CODE GOES HERE
```

Your explanation goes here!

3. Use the `max` function to compute the largest petal width in the data

```{r}
# YOUR CODE GOES HERE
```

Your explanation goes here!

4. Briefly describe the distribution of petal widths in the data using your answers from 1-3

```{r}
# YOUR CODE GOES HERE
```

Your explanation goes here!

5. Use the `hist` function to create a histogram of petal widths. Describe the distribution of the data.

```{r}
# YOUR CODE GOES HERE
```

Your explanation goes here!

# Book exercises

Complete ROS exercises 1.4, 1.5 and 2.7