---
title: "HW3"
author: "YOUR NAME"
date: "Today!"
format: pdf
---

```{r, include = F}
library(tidyverse)
```

For homework 3, complete the following exercises in ROS: Chapter 4: 4.1, 4.4

## 4.1 

```{r}
n <- 1000
n_t <- 500
n_c <- 500
r_t <- 0.5
r_c <- 0.4
# the average treatment effect (ATE) is the 
# difference in the outcome, response here
e <- r_t - r_c
# the standard error is s/sqrt(n)
# s for a binary is p*(1-p)
# so se for 1 is p*(1-p)/sqrt(n)
se_t <- (r_t * (1-r_t))/sqrt(n_t)
se_c <- (r_c * (1-r_c))/sqrt(n_c)
# the SE of a sum or difference of two variables is
# sqrt(se_1^2 + se_2^2)
se_e <- sqrt(se_t^2 + se_c^2)
```

The average treatment effect is a simple difference in mean response rates, so $0.5-0.4 = 0.1$. The standard error is the square root of the sum of squared standard errors (confusing!), which turns out to be `r se_e`

## 4.4

This one is a little tricky, we basically want to know what sample size we'll need to distinguish about a 0.1 difference. We could compute the n based on statistical power formulas, but that's messy. Instead, we'll just assume we need our SE for a single proportion to be about 0.05, so that 2 $\sigma$ is equal to about 0.1. 

To do this, we just need to start from the definition of s standard error for a proportion (see page 51-52):

$$se = \sqrt{\frac{p*(1-p)}{n}}$$

So to be sure a 30% shooter isn't really a 40% shooter, we'd need 

$$0.05 = \sqrt{\frac{0.3*(1-0.3)}{n}}$$
Solving for n, we square both sides and do some algebra

$$0.05^2 \times n = 0.21$$
$$n = 0.21/0.05^2$$

```{r}
0.21/0.05^2
```

We'll need about 84 shots to be sure.

## ggplot stuff

Then using the `mtcars` data built into R, complete the following using `ggplot`.

1. Create a histogram of fuel efficiency (mpg). Describe the distribution.

```{r}
ggplot(mtcars, aes(x = mpg)) + geom_histogram()
```

2. Create a histogram of vehicle weight (wt). Describe the distribution.

```{r}
ggplot(mtcars, aes(x = wt)) + geom_histogram()
```

3. Create a scatterplot of fuel efficiency by weight. Describe the joint distribution.

```{r}
ggplot(mtcars,
       aes(x = wt, y = mpg)) + 
  geom_point()
```

4. Include information on transmission type (automatic / manual; in the am variable) in the scatterplot. Describe the distribution.

```{r}
ggplot(mtcars,
       aes(x = wt, y = mpg, color = factor(am))) + 
  geom_point()
```

5. What can you conclude from these data visuals? Write 2-3 sentences to provide context. 


