---
title: "HW 1 Solutions"
author: "Frank Edwards"
date: "9/10/2019"
output: html_document
---

```{r echo = FALSE, message = FALSE}
library(tidyverse)
library(qss)
```

## Question 1

```{r}
data(turnout)
summary(turnout)
nrow(turnout)
ncol(turnout)
range(turnout$year)
```

There are `r nrow(turnout)` observations in this dataset, and `r ncol(turnout)` columns. The data contains observations every two years between 1980 and 2008, with the exception of 2006. The means, medians, ranges, and first and third quartiles of the data are presented above.

## Question 2

```{r}
turnout<-turnout %>% 
  mutate(turnout_vap = total / 
           (VAP + overseas),
         turnout_vep = total / (VEP))

turnout %>% 
  select(turnout_vap, turnout_vep)
```

As shown above, the turnout calculated by using the VAP rather than the VEP tends to be lower by several percentage points. The VAP is underestimating turnout relative to the VEP by an average of `r round(mean(turnout$turnout_vep - turnout$turnout_vap),3) * 100` percentage points between 1980 and 2008.

## Question 3

```{r}
turnout<-turnout %>% 
  mutate(ANES = ANES/100,
         ANES_diff_vap = turnout_vap - ANES,
         ANES_diff_vep = turnout_vep - ANES)

turnout_summary<-turnout %>% 
  summarise(mean_anes_diff_vap = mean(ANES_diff_vap),
            min_anes_diff_vap = min(ANES_diff_vap),
            max_anes_diff_vap = max(ANES_diff_vap),
            mean_anes_diff_vep = mean(ANES_diff_vep),
            min_anes_diff_vep = min(ANES_diff_vep),
            max_anes_diff_vep = max(ANES_diff_vep))

turnout_summary
```

The VEP yields results closer to the ANES estimated turnout on average, than does the VAP. The average difference between the VAP and ANES is `r round(turnout_summary$mean_anes_diff_vap,3)`, though this difference is as large as `r round(turnout_summary$min_anes_diff_vap,3)` and as small as `r round(turnout_summary$max_anes_diff_vap,3)` in some elections.

By contrast, the VEP differs from the ANES estimates on average by `r round(turnout_summary$mean_anes_diff_vep, 3)`, with a minimum difference of `r round(turnout_summary$min_anes_diff_vep,3)` and a maximum difference of `r round(turnout_summary$max_anes_diff_vep,3)`.

## Question 4

```{r}
### use the remainder from integer division to get presidential year
turnout<-turnout %>% 
  mutate(mod_4 = year%%4,
         pres_election = mod_4==2)

turnout_pres<-turnout %>% 
  filter(pres_election==TRUE) 

turnout_mid<-turnout %>% 
  filter(pres_election==FALSE)

mean(turnout_pres$ANES_diff_vep)
mean(turnout_mid$ANES_diff_vep)

```

On average, the bias of the VEP relative to the ANES is more extreme during mid-term elections. Between 1980 and 2008, the bias in the VEP derived estimate was  `r round(mean(turnout_pres$ANES_diff_vep) - mean(turnout_mid$ANES_diff_vep) ,3) * 100` percentage points lower in midterm elections versus general elections. 

## Question 5

```{r}
turnout_pre<-turnout %>% 
  filter(year<=median(year))
turnout_post<-turnout %>% 
  filter(year>median(year))

turnout_pre$ANES_diff_vep
turnout_post$ANES_diff_vep

mean(turnout_pre$ANES_diff_vep)
mean(turnout_post$ANES_diff_vep)
```

While there is some fluctuation, the ANES does appear to deviate more from VEP voting rate estimates in later years in this time series. The average bias in the post-1992 era is about 2 percentage points larger than the bias in the pre-1992 era.

## Question 6

```{r}
turnout_08<-turnout %>% 
  filter(year==2008) %>% 
  mutate(turnout_adj = (total - osvoters)/
           (VAP - felons - noncit))

turnout_08 %>% 
  select(ANES, turnout_vap, turnout_vep, 
         turnout_adj)
```

The ANES reported turnout rate is higher than all other turnout rates. Our adjusted figure based on a modified VAP comes closest to this fiugure, but is still 15 percentage points away from the ANES estimate. 

It appears that there is substantial bias in the ANES estimated voter turnout rates.