---
title: "Homework 1"
author: "YOUR NAME HERE"
date: "SUBMISSION DATE HERE"
output:
  html_document: default
  pdf_document: default
---

## 0.1 

Complete Swirl lesson INTRO2. 

Briefly describe one technique or concept in R that you are having trouble understanding. And don't worry! It is confusing and hard! 

THIS IS ALL NUTS AND I DON'T KNOW WHAT IS GOING ON!! But really, vectors are still confusing to me.

## Bias in voter turnout (book exercise 1.5.1)

You can obtain the file needed for this assignment either from the qss.student package, or from my web page:

https://raw.githubusercontent.com/f-edwards/intro_stats/master/hw/data/turnout.csv

Save the file as `turnout.csv` in the working directory that your homework file is in.

Surveys are frequently used to measure political behavior such as
voter turnout, but some researchers are concerned about the accuracy
of self-reports.  In particular, they worry about possible *social
desirability bias* where in post-election surveys, respondents who did
not vote in an election lie about not having voted because they may
feel that they should have voted.  Is such a bias present in the
American National Election Studies (ANES)?  The ANES is a nation-wide
survey that has been conducted for every election since 1948.  The
ANES conducts face-to-face interviews with a nationally representative
sample of adults.  The table below displays the names and descriptions
of variables in the `turnout.csv` data file. 



--------------------------------------------------------------------------------
 Name                 Description
 -------------------- ----------------------------------------------------------
 `year`               Election year
 
 `VEP`                Voting Eligible Population (in thousands) 
 
 `VAP`                Voting Age Population (in thousands) 
 
 `total`              Total ballots cast for highest office (in thousands) 
 
 `felons`             Total ineligible felons (in thousands) 
 
 `noncitizens`        Total non-citizens (in thousands) 
 
 `overseas`           Total eligible overseas voters (in thousands)
 
 `osvoters`           Total ballots counted by overseas voters (in thousands)
--------------------------------------------------------------------------------


## Question 1

Load the data into R and check the dimensions of the data.
  Also, obtain a summary of the data.  How many observations are
  there?  What is the range of years covered in this data set? 
  
  
```{r}
library(qss)
data(turnout)
summary(turnout)
nrow(turnout)
ncol(turnout)
range(turnout$year)
```

There are `r nrow(turnout)` observations in this dataset, and `r ncol(turnout)` columns. The data contains observations every two years between 1980 and 2008, with the exception of 2006. The means, medians, ranges, and first and third quartiles of the data are presented above.

## Question 2

Calculate the turnout rate based on the voting age population or
  VAP. Note that for this data set, we must add the total number of
  eligible overseas voters since the *VAP* variable does not
  include these individuals in the count. Next, calculate the turnout
  rate using the voting eligible population or VEP.  What difference
  do you observe?
  
```{r}
turnout$turnout_VAP <- turnout$total / (turnout$VAP + turnout$overseas)
turnout$turnout_VEP <- turnout$total / turnout$VEP

mean(turnout$turnout_VAP)
mean(turnout$turnout_VEP)
```

As shown above, the turnout calculated by using the VAP rather than the VEP tends to be lower by several percentage points. The VAP is underestimating turnout relative to the VEP by an average of `r round(mean(turnout$turnout_vep - turnout$turnout_vap),3) * 100` percentage points between 1980 and 2008.
  
## Question 3

Compute the difference between VAP and ANES estimates of turnout
  rate.  How big is the difference on average?  What is the range of
  the difference?  Conduct the same comparison for the VEP and ANES
  estimates of voter turnout.  Briefly comment on the results.
  
```{r}
turnout$ANES<-turnout$ANES/100

turnout$ANES_diff_vap <- turnout$turnout_VAP - turnout$ANES
turnout$ANES_diff_vep <- turnout$turnout_VEP - turnout$ANES

mean_anes_diff_vap <-mean(turnout$ANES_diff_vap)
min_anes_diff_vap <-min(turnout$ANES_diff_vap)
max_anes_diff_vap <-max(turnout$ANES_diff_vap)
mean_anes_diff_vep <-mean(turnout$ANES_diff_vep)
min_anes_diff_vep <-min(turnout$ANES_diff_vep)
max_anes_diff_vep <-max(turnout$ANES_diff_vep)

```

The VEP yields results closer to the ANES estimated turnout on average, than does the VAP. The average difference between the VAP and ANES is `r round(mean_anes_diff_vap,3)`, though this difference is as large as `r round(min_anes_diff_vap,3)` and as small as `r round(max_anes_diff_vap,3)` in some elections.

By contrast, the VEP differs from the ANES estimates on average by `r round(mean_anes_diff_vep, 3)`, with a minimum difference of `r round(min_anes_diff_vep,3)` and a maximum difference of `r round(max_anes_diff_vep,3)`.

## Question 4

Compare the VEP turnout rate with the ANES turnout rate
  separately for presidential elections and midterm elections.  Note
  that the data set excludes the year 2006. Does the bias of the ANES
  vary across election types?
  
```{r}
### every 4 years...
election_years<-seq(from = 1980, to = 2008, by = 4)

turnout$pres<-turnout$year%in%election_years

turnout_pres<-turnout[turnout$pres==T, ]

turnout_mid<-turnout[turnout$pres==F, ]

mean(turnout_pres$ANES_diff_vep)

mean(turnout_mid$ANES_diff_vep)
```

On average, the bias of the VEP relative to the ANES is more extreme during mid-term elections. Between 1980 and 2008, the bias in the VEP derived estimate was  `r round(mean(turnout_pres$ANES_diff_vep) - mean(turnout_mid$ANES_diff_vep) ,3) * 100` percentage points lower in midterm elections versus general elections. 


## Question 5

Divide the data into half by election years such that you subset
  the data into two periods.  Calculate the difference between the VEP
  turnout rate and the ANES turnout rate separately for each year
  within each period.  Has the bias of the ANES increased over time?
  
```{r}
turnout_pre<-turnout[turnout$year<=median(turnout$year), ]
turnout_post<-turnout[!(turnout$year<=median(turnout$year)), ]

turnout_pre$ANES_diff_vep
turnout_post$ANES_diff_vep

mean(turnout_pre$ANES_diff_vep)
mean(turnout_post$ANES_diff_vep)
```

While there is some fluctuation, the ANES does appear to deviate more from VEP voting rate estimates in later years in this time series. The average bias in the post-1992 era is about 2 percentage points larger than the bias in the pre-1992 era.

## Question 6

The ANES does not interview overseas voters and
  prisoners. Calculate an adjustment to the 2008 VAP turnout
  rate. Begin by subtracting the total number of ineligible felons and
  non-citizens from the VAP to calculate an adjusted VAP. Next,
  calculate an adjusted VAP turnout rate, taking care to subtract the
  number of overseas ballots counted from the total ballots in 2008.
  Compare the adjusted VAP turnout with the unadjusted VAP, VEP, and
  the ANES turnout rate. Briefly discuss the results.
  

```{r}
turnout_08<-turnout[turnout$year==2008, ]
turnout_08$turnout_adj<- (turnout_08$total - turnout_08$osvoters) / 
  (turnout_08$VAP - turnout_08$felons - turnout_08$noncit)

turnout_08$turnout_adj; turnout_08$turnout_VAP; turnout_08$turnout_VEP
```

The ANES reported turnout rate is higher than all other turnout rates. Our adjusted figure based on a modified VAP comes closest to this fiugure, but is still 15 percentage points away from the ANES estimate. 

It appears that there is substantial bias in the ANES estimated voter turnout rates.